---
title: "GDP components over time and among countries"
author: "Nithish Kumar"
date: "`r Sys.Date()`"
tags: ["Markdown", "GDP"]
output:
  html_document:
    theme: flatly
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show
editor_options: 
  markdown: 
    wrap: 72
---

```{r, setup, echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  echo = FALSE,
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)
```

```{r load-libraries, echo=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(GGally)
library(readxl)
library(here)
library(skimr)
library(janitor)
library(broom)
library(tidyquant)
library(infer)
library(openintro)
```

Let's look at the GDP data from the [United Nations' National Accounts
Main Aggregates Database](https://unstats.un.org/unsd/snaama/Downloads),
which contains estimates of total GDP and its components for all
countries from 1970 to today. At the risk of oversimplifying things, the
main components of gross domestic product, GDP are personal consumption
(C), business investment (I), government spending (G) and net exports
(exports - imports).

We will look at how GDP and its components have changed over time, and
compare different countries and how much each component contributes to
that country's GDP.

```{r read_GDP_data, echo=FALSE, results='hide'}

UN_GDP_data  <-  read_excel(here::here("content/posts/GDP_components_among_countries/data", "Download-GDPconstant-USD-countries.xls"), # Excel filename
                sheet="Download-GDPconstant-USD-countr", # Sheet name
                skip=2) # Number of rows to skip

```

First Let's load the data and tidy it. Let's compare the GDP changes for
United States, Germany and India.

```{r reshape_GDP_data, echo=FALSE, results='hide'}

tidy_GDP_data  <- UN_GDP_data %>% 
  # Pivot data into longer format
  pivot_longer(cols = "1970":"2017", 
               names_to = "year", 
               values_to = "value") %>% 
  # Express figures in billions
  mutate(value = value / 10^9) %>% 
  # Rename indicators
  mutate(IndicatorName = case_when(
    IndicatorName == "Exports of goods and services" ~ "Exports",
    IndicatorName == "Imports of goods and services" ~ "Imports",
    IndicatorName == "Gross capital formation" ~ "Gross_capital_formation",
    IndicatorName == "General government final consumption expenditure" ~ "Government_expenditure",
    IndicatorName == "Household consumption expenditure (including Non-profit institutions serving households)" ~  "Household_expenditure",
    IndicatorName == "Imports of goods and services" ~ "Imports",
    IndicatorName == "Gross Domestic Product (GDP)" ~ "GDP_from_raw_data",
    TRUE ~ IndicatorName)
  )

# Let us compare GDP components for these 3 countries
country_list <- c("United States","India", "Germany")

```

Firstly, let's see how each GDP components has changed for these 3
countries since 1970.

```{r gdp1_plot, out.width="100%"}
tidy_GDP_data %>% 
  # Only consider selected countries
  filter(Country %in% country_list) %>% 
  # Only consider selected indicator names
  filter(IndicatorName %in% c("Gross_capital_formation", "Exports", "Government_expenditure", "Household_expenditure", "Imports")) %>% 
  # Plot the graph, group according to IndicatorName to receive one line per indicator
  ggplot(aes(x = year, y = value, group = IndicatorName, color = IndicatorName)) + 
    geom_line() + 
    facet_wrap(~ Country) + 
  theme_bw() + 
  labs(title = "GDP components over time", 
       subtitle = "In constant 2010 USD", 
       y = "Billion USD", 
       x = NULL)

```

Now, let's compare the GDP and its breakdown at constant 2010 prices in
US Dollars for United States, India and Germany.

```{r gdp2_plot, out.width="100%"}
tidy_GDP_data2 <- tidy_GDP_data %>% 
  filter(IndicatorName %in% c("Gross_capital_formation", "Exports", "Government_expenditure", "Household_expenditure", "Imports", "GDP_from_raw_data")) %>% 
  filter(Country %in% country_list) %>% 
  # Pivot wider to calculate Net Exports
  pivot_wider(names_from = IndicatorName, 
              values_from = value) %>% 
  mutate(Net_exports = Exports - Imports) %>% 
  # Deselect Exports and Imports because they are no longer needed
  select(-Exports, -Imports) %>% 
  # Calculate the GDP based on its components
  mutate(GDP_calculated = Gross_capital_formation + Net_exports + Government_expenditure + Household_expenditure)

# Find out % difference between calculated GDP and GDP from raw data
tidy_GDP_data2 <- tidy_GDP_data2 %>% 
  mutate(sum_GDP_calculated = sum(GDP_calculated))

# Continue with plot
tidy_GDP_data3 <- tidy_GDP_data2 %>% 
  # Calculate GDP component values as percentages of total GDP
  mutate(Household_expenditure = Household_expenditure / GDP_calculated, 
         Government_expenditure = Government_expenditure / GDP_calculated, 
         Gross_capital_formation = Gross_capital_formation / GDP_calculated, 
         Net_exports = Net_exports / GDP_calculated) %>% 
  select(-GDP_from_raw_data, -GDP_calculated) %>% 
  # Pivot longer to facilitate plotting
  pivot_longer(cols = Household_expenditure:Net_exports, 
               names_to = "indicator", 
               values_to = "value")

# Plot the data
ggplot(tidy_GDP_data3, aes(x = year, y = value, group = indicator, color = indicator)) + 
  geom_line() + 
  facet_wrap(~Country) + 
  theme_bw() + 
  # Set labels to percentages
  scale_y_continuous(labels = scales::percent) + 
  labs(title = "GDP and its breakdown at constant 2010 prices in US Dollars", 
       y = "Proportion", 
       x = NULL)

```

These are our observations from keenly observing these graphs for United
States, Germany and India.

1.  In Germany, the only GDP component that has been growing for the
    past 20 years is net exports. This indicates that German economic
    growth is fueled by exporting more than importing. Government
    expenditure is relatively stable. This indicates that the government
    isn't investing heavily in the moment, which is sadly true and which
    explains why Germany is very backward in terms of internet access
    and future technology adoption (like 5G networks). Gross capital
    formation (also called "investment") has been going down lightly for
    the past 50 years, indicating that German businesses don't invest as
    heavily as they used to. Relative German household expenditure has
    been stable. This makes sense when considering that the German
    population isn't growing, it is rather shrinking and only an
    immigration influx has been stablilizing it.

2.  The chart looks very different for India. There, relative household
    expenditure has dropped from around 70% in 1970 to around 55%. In
    the same time period, gross capital formation has been going up by
    15%. This indicates that Indian companies are investing heavily at
    the moment. Furthermore, government expenditure has been mostly
    stable at around 10%, which is half as much as Germany. The
    frugality of the Indian government probably stems from the fact that
    India is not yet as highly developed as Germany. Interestingly, net
    exports has been mostly negative in the past 50 years. This
    indicates that India is reliant on imports from other countries
    because they import more than they export.

3.  In the US, the component that is growing the most is household
    expenditure, now at about 70%. In Germany, this is currently only
    55%. This indicates that the US economy focuses more on producing
    for their own population compared to Germany focusing a bit more on
    exporting goods. This makes sense as the US market is extremely
    strong. In addition, the US have negative net exports, currently at
    around -5%. Relative government expenditure has been decreasing,
    from around 25% in 1970 to around 15% in 2016. In Germany, this
    value is at around 20% in 2016. This indicates that the US isn't
    investing heavily at the moment, growth rather stems from household
    expenditure.
